Refinement for open automata

### 2021-06-03 2021-Refinement/report.tex
Fixed proof of lemma 1

% transitions between sections
% TODO: CITER LES CONFERENCES
% change deadlock reduction into deadlock prevention?

@@

\section{Introduction}
The open automata model is used to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
Verification is the process of checking that an implementation satisfies a specification.
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify BIP and GCM \cite{qin:01823507, ameurboulifa:01526055}.
However equivalences relations are not always sufficient to perform verification, refinement relations are also used.
For instance an example of open pNet where equivalence is not sufficient was encountered in ...\marginpar{Paper where the motivating example came up}.

@@

\section{Notations}\label{sec:notations}
Notations will be defined with the operator \(\defnotation\) and names are given with the operator \(\defobject\) as follows:
\begin{align*}
	notation\_with\_variables & \defnotation notated\_object\_using\_the\_variables \\
	name & \defobject fully\_defined\_mathematical\_object
\end{align*}

@@

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar to the previous one, as proved in \marginpar{probably forte, check that}.
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds.
The variables local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
In a run, when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

@@

Composition is a complex process that generates big automata with complex and potentially simplifiable transitions.
However simplification is a hard problem that is not yet automatised on open automata, so from this point the guards and transitions will always be simplified by hand without further explaination.
An example of composition can be found in Appendix \ref{apx:composition}.
\begin{prop}[Parallel composition]
Let \(A_1\), \(A_2\), \(A_3\) be three open automata with respectively \(J_1\), \(J_2\) and \(J_3\) as hole names.
We have
\begin{align*}
	A_1\subst{A_2}{j}\subst{A_3}{j'} & = A_1\subst{A_3}{j'}\subst{A_2}{j} & j, j' & \in J_1 \\
	\mpar{A_1\subst{A_2}{j_1}}\subst{A_3}{j_2} & = A_1\subst{A_2\subst{A_3}{j_2}}{j_1} & \mpar{j_1, j_2} & \in J_1 \times J_2
\end{align*}
Where the equality is modulo isomorphism (state renaming, associativity, commutativity).
\end{prop}
This property is induced by the composition on open pNets.
Intuitively it holds because the cartesian product of states, the union of sets, the union of maps and the conjunction on formulas (in the guards) are commutative and associative.
This allows us to define the simultaneous composition \(A\psubst{A_j^{j \in J'}}\) as the result of any order of (sequential) composition where each \(A_j\) is composed in the hole \(j\). % psubst to pcomp macro [[]]

\section{Properties of a refinement relation for Open Automata}\label{sec:proofelts}
There are several properties we may expect from a refinement relation.
Depending on these properties, different kinds of refinement can be used.
For example if we are interested in producing the same sequences of actions as another automaton we may want to use trace set inclusion as a refinement.
Here the main expected properties are related to composition and action refinement. % LATER: explain action refinement here or remove if not tackled
Because of composition, a relation as strong as simulation must be used as observed in the articles about FH-bisimulation \cite{henrio:01055091}.

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
Generally the second is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property, and only talking about \(R\) and \(P\) is sufficient.
However later on in this article we will define more complex relations on automata so we clarify the distinction now.
A relation of second kind will be called ``relation on automata" whereas a relation of the first kind is called ``relation on configurations" as defined below.
In the work on FH-bisimulation\marginpar{re-cite FH-bisim as near as the previous citation?} such a relation \(R\) was defined as triples in \(S_1 \times S_2 \times \formulas\) where for \(\mpar{s_1, s_2} \in S_1 \times S_2\) there is a unique \(f \in \formulas\) such that \(\mpar{s_1, s_2, f} \in R\).
This is classically the definition of a function from \(S_1 \times S_2\) to \(\formulas\) so we will define it as such.
\begin{defi}[Relation on configurations]
A relation on configurations of two open automata \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).

@@

The important properties we consider for the relation on states are being a simulation and the prevention of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\psubst{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

@@

\section{Proof of equivalent definitions}\label{apx:lemeqd}

@@

	\(R\) satisfies \eqref{eq:drWD} with the current value of \(\mpar{s_1, s_2}\) and \(\sigma \uplus \nu\) completed with dummy values for the other variables of \(\biguplus_{t_2 \in \fOT{s_2}} \fvars{\fguard{t_2}}\) so we have:

@@

	The \(\vdash\) hides an implicit \(\exists\) that we use to get values to the variables in \(\biguplus_{t_1 \in \fOT{s_1}} \fvars{\fguard{t_1}}\) in a valuation \(\mu\).

@@

	At this point we have a \(t_1\) and a \(t_2\) so we may think that the proof is about to be finished because the only property left to prove is:

@@

	However we cannot yet prove it because we don't know whether \(t_2\) and \(t_1\) do match.
	In order to get a \(t_2\) matching \(t_1\) we use will use \eqref{eq:drID}.
	\eqref{eq:drID} with current value of \(\mpar{s_1, s_2}\), \(t_1\) and \(\sigma \uplus \mu\) as a valuation of \(V_1 \uplus V_2 \uplus \fvars{t_1}\) gives:

@@

	The witnesses for \goal{1} are \(t'_2\) and \(t_1\), and the rest is proved with a combination of \hyp{target} and \hyp{3'}.

@@

We have the hypotheses on \(R_{12}\) and \(R_{23}\), then the current goal: % TODO HERE: get s_2 by decomposing R_{13}

@@

	\sigma_2 & : V_2 \to \values &
	\sigma_1 \uplus \sigma_2 \uplus \sigma_3 & \vdash R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3} \hyp{7'} \\
	&& \sigma_1 \uplus \sigma_3 & \vdash \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} \hyp{7"}

>>>>

% transitions between sections
% TODO: CITER LES CONFERENCES

@@

\section{Introduction}
The open automata model is used to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
Verification is the process of checking that an implementation satisfies a specification.
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify BIP and GCM \cite{qin:01823507, ameurboulifa:01526055}.
However equivalences relations are not always sufficient to perform verification, refinement relations are also used.
For instance an example of open pNet where equivalence is not sufficient was encountered in an article about algorithms fo weak bisimulation on open automata \cite{wang:hal-03126313}.

@@

\section{Notations}\label{sec:notations}
Notations will be defined with the operator \(\defnotation\) and names are given with the operator \(\defobject\) as follows:
\begin{align*}
	\mathit{notation\_with\_variables} & \defnotation \mathit{notated\_object\_using\_the\_variables} \\
	\mathit{name} & \defobject \mathit{fully\_defined\_mathematical\_object}
\end{align*}

@@

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar to the previous one \cite{henrio:01299562}.
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds.
The variables local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
In a run, when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

@@

Composition is a complex process that generates big automata with complex and potentially simplifiable transitions.
However simplification is a hard problem that is not yet automatised on open automata, so from this point the guards and transitions will always be simplified by hand without further explaination.
An example of composition can be found in Appendix \ref{apx:composition}.
\begin{prop}[Parallel composition]
Let \(A_1\), \(A_2\), \(A_3\) be three open automata with respectively \(J_1\), \(J_2\) and \(J_3\) as hole names.
We have
\begin{align*}
	A_1\subst{A_2}{j}\subst{A_3}{j'} & = A_1\subst{A_3}{j'}\subst{A_2}{j} & j, j' & \in J_1 \\
	\mpar{A_1\subst{A_2}{j_1}}\subst{A_3}{j_2} & = A_1\subst{A_2\subst{A_3}{j_2}}{j_1} & \mpar{j_1, j_2} & \in J_1 \times J_2
\end{align*}
Where the equality is modulo isomorphism (state renaming, associativity, commutativity).
\end{prop}
This property is induced by the composition on open pNets.
Intuitively it holds because the cartesian product of states, the union of sets, the union of maps and the conjunction on formulas (in the guards) are commutative and associative.
This allows us to define the simultaneous composition \(A\mdbrk{A_j^{j \in J'}}\) as the result of any order of (sequential) composition where each \(A_j\) is composed in the hole \(j\).
Even if an automaton is composed in two distinct holes it does not share its variables, each composition makes the cartesian product of states and a disjoint union of the variables.
In practise, variables may have to be renamed.

\section{Properties of a refinement relation for Open Automata}\label{sec:proofelts}
There are several properties we may expect from a refinement relation.
Depending on these properties, different kinds of refinement can be used.
For example if we are interested in producing the same sequences of actions as another automaton we may want to use trace set inclusion as a refinement.
Here the main expected properties are related to composition and action refinement. % LATER: explain action refinement here or remove if not tackled
Because of composition, a relation as strong as simulation must be used.

@@

Note that in the previous paragraph there are two relations:
The first is between states of two specific automata (\(R \subseteq S_1 \times S_2\)), the second is between automata (\(A_1 \leq A_2\)).
Generally the second is implicit because its definition is always \(\exists R \subseteq S_1 \times S_2, P\mpar{R}\) with \(P\) a property, and only talking about \(R\) and \(P\) is sufficient.
However later on in this article we will define more complex relations on automata so we clarify the distinction now.
A relation of second kind will be called ``relation on automata" whereas a relation of the first kind is called ``relation on configurations" as defined below.
In the work on FH-bisimulation \cite{henrio:01055091} such a relation \(R\) was defined as triples in \(S_1 \times S_2 \times \formulas\) where for \(\mpar{s_1, s_2} \in S_1 \times S_2\) there is a unique \(f \in \formulas\) such that \(\mpar{s_1, s_2, f} \in R\).
This is classically the definition of a function from \(S_1 \times S_2\) to \(\formulas\) so we will define it as such.
\begin{defi}[Relation on configurations]
A relation on configurations of two open automata \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).

@@

The important properties we consider for the relation on states are being a simulation and the prevention of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This definition translated to open automata the classical definition of simulation \marginpar{cite: Fernandez-Mounier-91-b}.
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

@@

\section{Proof of equivalent definitions}\label{apx:lemeqd} % TODO: disjunction branch, say where we get the value

@@

	\(R\) satisfies \eqref{eq:drWD} with the current value of \(\mpar{s_1, s_2}\) and \(\sigma \uplus \nu\) completed with dummy values for the other variables of \(\biguplus_{t_2 \in \fOT{s_2}} \fvars{t_2}\) so we have:

@@

	The \(\vdash\) hides an implicit \(\exists\) that we use to get values to the variables in \(\biguplus_{t_1 \in \fOT{s_1}} \fvars{t_1}\) in a valuation \(\mu\).

@@

	At this point we have a certain \(t_1\) and \(t_2\) so we may be tempted to use them to prove the goal:

@@

	However we cannot yet prove it because we don't know whether \(t_2\) and \(t_1\) match.
	In order to get a \(t_2\) matching \(t_1\) we will use the fact that \(R\) is a simulation.
		\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
		\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2} \hyp{sim}
	\hyp{sim} with current value of \(\mpar{s_1, s_2}\), \(t_1\) and \(\sigma \uplus \mu\) as a valuation of \(V_1 \uplus V_2 \uplus \fvars{t_1}\) gives:

@@

	The witnesses for \goal{1} are \(t'_2\) and \(t_1\), and the rest is proved with a combination of \hyp{4} and \hyp{3'}.

@@

We have the hypotheses on \(R_{12}\) and \(R_{23}\) for a specific \(s_2\), then the current goal:

@@

	\sigma_1 \uplus \sigma_2 \uplus \sigma_3 \vdash R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3} \hyp{7'} \\
	\sigma_1 \uplus \sigma_3 \vdash \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} \hyp{7"}
Hypotheses \hyp{5} and \hyp{6} are considered with the obtained state \(s_2\).
